Data Exploration, Quality Control, and Statistical Analysis of ChIP-exo/nexus Experiments

### 2016-07-07 paper/chipexo_QC_vs2.tex
edited the results part and figures that correspond to the first results part

The Strand Cross-Correlation (SCC), introduced by \cite{strandcc}, is the most commonly used quality metric in assessing ChIP-seq enrichment quality. It aims to quantify  how well the reads mapped to each strand are clustered around the locations of the protein-DNA interaction sites by calculating the Pearson correlation among forward and backward strands reads by shifting them across a range that covers both the read length of the experiment and the expected average fragment length. Typical SCC profiles exhibit two local maxima: at the average fragment length and the read length. In high quality experiments with clear ChIP enrichment, the average fragment length maximum coincides with the global maximum. In an idealized ChIP-exo experiment where the DNA fragments are digested to the boundaries of the protein-DNA interaction sites, we would expect the SCC profile to maximize at the motif length indicating clustering of the forward and reverse strand reads around the binding site. Figure~\ref{fig:1}D displays the SCC curves for the CTCF HeLa samples where the ChIP-exo curve shows local maxima at the motif and read lengths, while the SE ChIP-seq curves have a local maxima at the read length and a global maxima at the average fragment length. SCC profiles for other samples are available in Supplementary Figures \ref{sfig:scc1} to \ref{sfig:scc8}. The read length and motif length maxima are often in close proximity of each other; as a result, this renders QC metrics such as the Normalized Strand Cross-Correlation (NSC) or the Relative Strand Cross-Correlation (RSC) harder to interpret; however, the profile itself seems informative about the enrichment signal in ChIP-exo/nexus experiments.

% \SK{this seems true at least based on my initial look at the plots, but we should be cautious in reporting this}. 
% \RW{i agree, and additionally Carroll et al, 2014 kinda agrees with us that the SCC-related measure may not be useful since the fragment length coincides with read length}

% http://journal.frontiersin.org/article/10.3389/fgene.2014.00075/full#h5 the paragraph is a little bit above conclusions

% We continued our exploration by investigating whether the current state-of-the-art QC pipelines for ChIP-Seq are suitable for ChIP-exo %and ChIP-nexus. In Tables \ref{tab:qc_sig} and \ref{tab:qc} we calculated a collection of the commonly used ChIP-Seq QC metrics using the ChIP-exo and ChIP-nexus experiments instead: Normalized Strand Cross-Correlation (NSC), Relative Strand Cross-Correlation and PCR Bottleneck Coefficient (PBC) defined as in \url{https://genome.ucsc.edu/ENCODE/qualityMetrics.html}\nocite{encode_qc}.

% considerable amounts of reads are being mapped to specific positions.

% because of the exonuclease digestion, several reads that their $5\prime$ end was digested to the same position before the amplification step.

%, which by following the ChIP-Seq guidelines it would indicate that those experiment show severe bottlenecking problems, which may incorrectly suggest that the positions with a large amount of aligned reads are being caused by PCR amplification rather than observed genomic signal.

%, and usually it is expected to observed two local maxima, one when the profiles are shifted by the average read length and another when the profiles are shifted by the unobserved fragment length. In a high quality ChIP-Seq dataset the last one is also the SCC global maxima.

@@

We first present the overall pipeline and then discuss individual
components with a case study using ChIP-exo data of FoxA1 from
\cite{exoillumina} and ChIP-nexus data from \cite{chipnexus}.
Figure~\ref{fig:2} summarizes the 4-step pipeline.  Given
aligned reads from a ChIP-exo/nexus sample, the first step partitions
the reference genome into islands by keeping the non-digested ChIP-exo
regions.  In step 2, the total number of extended reads (by the
experiment's read length) \SK{check: extended or not? What do we
  extend to in ChIP-exo?} \RW{right now, we don't extend. we use the
  coverage plot with the experiment's read length to partition the
  coverage into the regions. In the package I left it as a parameter
  but I don't is going to change a lot. For example, the ChIP-exo
  datasets read length is ~45 and it's regions seems to be composed by
  scattered reads in both sides. So, I think that using a higher value
  will only made the regions wider but will not increase the number of
  unique position or the depth by a lot} overlapping each island
($D_i$) and the number of unique island positions with at least one
aligned read re recorded ($(U_i)$. Then, three summary statistics
$\text{ARC}_i$, $\text{URC}_i$, and $\text{FSR}_i$ are computed for
each region $i$.  $\text{ARC}_i$ denotes the \textit{average read
  coefficient} and is defined as the ratio of the \# of reads in
island $i$ ($D_i$) to the width of the island $i$ ($W_i$);
$\text{URC}_i$, \textit{unique read coefficient}, quantifies the
inverse of the effective coverage and is defined as the ratio of the
\# of genomic positions with at least one aligned read within island
$i$ ($U_i$) to the \# of reads in island $i$ ($D_i$); and
$\text{FSR}_i$ denotes the proportion of forward strand reads.
%\begin{align*}
%  \text{ARC}_i &= \dfrac{\text{\#  of reads in island $i$}}{\text{Width of the island $i$}}, \\
%  \text{URC}_i & = \dfrac{\text{\# of genomic positions with at least
%      one aligned read within island $i$}}{\text{\# of reads in
%      island %$i$}}, \\
%      \text{FSR}_i &= \dfrac{\text{\# of forward strand reads in
%          island $i$}}{\text{\# of reads in island $i$}}.
%\end{align*}
Step 3 of the pipeline generates several diagnostic plots aimed at
quantifying ChIP enrichment and strand imbalance and step 4 generates
quantitative summaries of these diagnostic plots.

Figure~\ref{fig:2}A presents the typical behavior of the URC vs. ARC
plot for a high quality ChIP-exo sample. In general, the plot depicts
two strong arms.  High URC values correspond to regions with reads
concentrated on a small number of positions.  The left arm, with low
ARC and varying URC values, corresponds to background islands, regions
that are usually composed of scattered reads that were not digested
during the exonuclease step. The right arm where the URC decreases as
the ARC increases corresponds to regions that are usually ChIP
enriched. \RW{Finally we quantify the shape of the ARC vs. URC plot by
  the use of two estimated parameters: $\beta_1$ which
  represents the average number of reads aligned to the unique
  positions in large depth regions and $\beta_2$ which
  represents the overall change in depth as the width varies across a
  large set of regions.}  \RW{Figures~\ref{fig:2}B and \ref{fig:2}C
  presents the typical behavior of the Region Composition and Forward
  Strand Ratio distribution plots respectively, both quantify the
  strand imbalance as part of the QC pipeline. The former depicts how
  quickly the islands exclusively composed by fragments on a single
  strand are filtered out as islands with higher depths are
  observed. In a high quality sample, the proportion of islands with
  reads from only one strand is expected to decrease rapidly as we
  consider higher depth regions. In contrast, this proportion remains
  approximately constant in lower quality samples. The latter plot
  illustrates how quickly the quantiles of the FSR approaches to 0.5,
  the expected FSR value in high quality samples. Even though not
  every region in a ChIP-exo experiment is perfectly balanced, the
  most enriched regions are expected to be composed by approximately
  the same quantity of reads in both strands.}

%and as the URCR decreases the library complexity does it as well, on
%the other hand 
% \SK{Not sure what this next sentence is referring to: Finally we
%   quantify the relationship between library complexity by the use of
%   two indexes that represent the change in the number of unique
%   positions per regions and the change in of the width of a regions as
%   the depth changes.}

%median, since in a high quality sample it is
%expected for the median to be approximately 0.5 and the enriched
%regions are going to be composed by fragments sequenced from both
%strands.

\subsubsection*{Application and validation of \pname{} with the FoxA1
  ChIP-exo dataset}
% \SK{We should present the whole QC results and end with validation -  panels from the two figures should be switched.}

We next use Fox A1 ChIP-exo datasets, with three biological replicates
at comparable sequencing depths from mouse liver cells, to illustrate
the QC pipeline.  Figure~\ref{fig:3}A presents URC vs. ARC plots
for all three replicates. The first and third replicates exhibit a
defined decreasing trend in URC as the ARC increases. This indicates
that these samples exhibit a higher ChIP enrichment than the second
replicate. On the other hand, the overall URC level from the first two
replicates is higher than that of the third replicate level,
elucidating that the libraries for the first two replicates are more
complex than that of the third replicate.

%As we discussed above,  background fragments are often digested by
%the exonuclease enzyme in ChIP-exo experiments; therefore, the balance between the enrichment
%and library complexity of an experiment is one of the key factors determining
%the sample's data quality.
%Using the Fox A1 in mouse liver cell lines generated by Serandour et
%al., 2013\nocite{exoillumina} and these two quantities, we explored
%the relationship between library complexity and experiment
%enrichment.

@@

%The strand imbalance assessment is based in the observation that the
%enriched regions usually are composed of a higher quantity of reads,
%therefore we examined the FSR as the regions with lower depth are
%being filtered out. This indicator is of particular importance, as it
%evaluates the ``peak pair'' assumption that the original ChIP-exo
%paper suggested and multiple ChIP-exo data analysis methods rely
%on. For every ChIP-exo experiment, we calculated the global FSR and
%noticed that for all experiments is roughly $0.5$, which means there
%are approximately the same amount of reads in both strands.

\SK{I have a hard time interpreting B and C specifically, need to
  include something about their implications for replicates 2 and 3.}

Figure~\ref{fig:3}B and \ref{fig:3}C display the Read Composition and
FSR distribution plots, which highlight specific problems with
replicates 2 and 3. \RW{In Figure~\ref{fig:3}B, we can observe
  decreasing trends in the proportions of regions formed by fragments
  in one exclusive strand. In a high quality experiment its is
  expected to observe an exponential decay in the proportion of single
  stranded regions, while in lower quality experiments the trend may
  be linear of even constant (Supplement
  Figure~\ref{sfig:qcsamp3}). In the FSR distribution plot, both
  replicate 2 and 3 FSR's distributions are more spread around their
  respective medians. The approximation of the 0.1 and 0..9 quantiles
  to the median indicated the aforementioned lower enrichment in the
  second replicate and the low complexity in the third one.
 }
 % In particular, \RW{both replicate 2 and replicate 3 FSR's
 %  distributions are more spread around the median. The approximation
 %  of the 0.1 and 0.9 quantiles to the median is indicative of the
 %  aforementioned lower enrichment of the second replicate, since there
 %  are less regions with a higher concentration of reads, this makes
 %  the FSR's distribution to be dominated by non-enriched regions. On
 %  the other hand, the slow approximation of the quantiles to the FSR
 %  median in the third replicate is indicative of low library
 %  complexity. In Figure~\ref{fig:3}C, we can observe an decreasing
 %  trend in the proportion of regions formed by fragments in one
 %  exclusive strand. In a high quality experiment it is expected to
 %  observe an exponential trend, while in lower quality experiments the
 %  trend may be linear or even constant (Supplement
 %  Figure~\ref{sfig:qcsamp3}).}

%In order to study the strand imbalance of the FoxA1 factor ChIP-exo
%experiments in mouse liver generated by Serandour et al.,
%2013\nocite{exoillumina} we considered the visualizations shown in
%Figure \ref{fig:strand}: Figure \ref{fig:strand}A presents the FSR's
%behavior as the lower depth regions are being filtered out, while
%Figure \ref{fig:strand}B) shows the proportion of regions that are
%composed by reads from both strands against the regions formed by
%reads in exclusively one strand. In a high quality dataset, it is
%expected for all quantiles to quickly converging towards the median
%(in panel A) or the regions formed by reads in one strand (either
%forward or backward) to be formed by few fragments (in panel
%B). 
%Additionally, it is worth mentioning that the background in a ChIP-exo
%experiment is expected to be formed by the undigested ChIP fragments
%that are not bound to a protein, hence it is for the reads in both
%strand to be imbalanced in a background region.

 Overall, we conclude that replicate 1 is higher quality than both of
 replicates 2 and 3.  We validate this observation with a motif
 analysis on the candidate binding regions identified from these
 replicates \SK{Could we say binding events? or is the peak analysis
   on the whole binding region, i.e., peak? If so, we may want o
   change "candidate regions" to peaks?} \RW{I used a set of high
   quality regions, i.e. using the ChIP-exo pipeline I filtered them
   by npos, FSR and depth. Figure~\ref{sfig:venn} show the overlap
   between these regions and the peaks (roughly speaking these regions
   are almost a proper subset of the peaks)}. Figure~\ref{fig:4}A
 summarizes the total numbers of regions identified from each
 replicate. The lower number of enriched regions from replicate 2 is
 consistent with the lower ChIP enrichment pattern in the $UCR$
 vs. $ARC$ diagnostic plot.  Scanning of these regions for the
 occurrence of FoxA1 sequence motif with the FIMO tool \cite{fimo}
 indicates that the first replicate outperforms the other two in terms
 of percentage of candidate regions with the FoxA1 motif
 (Figure~\ref{fig:4}B).

%Considering the regions and summary statistics we obtained a set of
%candidate sites for each replicate, the total number of candidate
%sites is shown in Figure \ref{fig:enrich}B.
%For this sites, we
%extracted the sequence and searched for the FoxA1 motif using
%\textbf{FIMO} \cite{fimo}. Figure \ref{fig:enrich}C shows the fraction
%of candidate site where a motif was found. The first replicate
%outperforms the rest in both number of candidate sites and proportion
%of sites with motif; on the other hand we can see that the second
%replicate candidate sites are more likely to contain the motif
%sequence, but there more candidate sites in the third replicate than
%in the second. 
Furthermore, Figure~\ref{fig:4}C displays the average normalized
read coverage around the actual motif locations in the candidate
binding regions.  These coverage plots reveal that the ChIP signal is
more defined for the first and third replicates than the second one,
indicating overall strength of the ChIP enrichment in these samples
compared to the second replicate.
% however,  overall, the signal pattern is reproducible in all of the three biological replicates.
Figure~\ref{fig:4}D further highlights the overall quality of the
identified motif sequences for each replicate and suggest that
libraries with high library complexity (replicates 1 and \RW{2})
capture binding sites with better motif
matches. \RW{Figure~\ref{fig:4}E compares the top FIMO scores among
  the three replicates, not-surprisingly confirming that the first
  explicate exhibits the highest quality. Finally, Figure~\ref{fig:4}F
  compares the Strand Imbalance by considering the regions overlapping
  with their respective sets of ChIP-exo peaks. Not surprisingly, for
  the replicate 2 we can see that the Imbalance distribution differ
  among both classes, which is caused by the aforementioned lower
  enrichment of that sample; for the remaining replicates we can see
  that the Imbalance is more similar when regions composed by more
  reads are considered. This suggests that enriched regions are more
  likely to be balanced, therefore this suggest that the ``peak-pair''
  assumption may not hold in non-enriched regions.}

 % Hence, in a low enriched sample the
 %  ``peak-pair'' assumption does not hold globally.}
% \SK{I also don't follow the discussion for panel C: Figure
% \ref{fig:4}C compares the strand imbalance when ChIP-exo islands 
% overlap with high quality peaks. It is noticeable that for regions
% composed by a large amount of reads, it is harder to distinguish their
% peaks by considering only the strand imbalance, hence in a high
% quality ChIP-exo experiment the background is expected to show a
% higher imbalance that the enriched regions. In conclusion, Figure
% \ref{fig:4} shows that the global FSR does not accurately
% represent a ChIP-exo experiment's strand imbalance locally, hence the
% ``peak pair'' assumption does not completely hold in every ChIP-exo
% enriched region.}
% \SK{Worth thinking a bit more about the interpretation.}

%In summary,  the diagnostics plots of the QC pipeline highlights that  observing a defined
%decreasing trend in $\mbox{URCR}$ as $\mbox{ARC}$ increases implies
%ChIP enrichment and multiple samples can be compared with respect to the degree of this trend. 
%Second, for experiments that are not deeply
%sequenced, low $\mbox{URCR}$ values represent low complexity. 
%decreases the amount of sites of motif found but it doesn't modify the
%read distribution around the motif. Finally, libraries with higher
%complexity returns sequences with a better motif match. \SK{We do not have enough analysis with motif validation to conclude these.}

\subsubsection*{High sequencing depth may confound low-complexity
  library issues}
We next evaluated every sample listed in Tables~\ref{tab:qc} and
\ref{tab:qc_sig} with the \pname{} QC pipeline (Supplementary
Figures~\ref{sfig:qc1} to \ref{sfig:qc11}). A key, albeit not
surprising, observation from large scale analysis is that the URC
vs. ARC plots typically display the three patterns captured in the
FoxA1 study. We will refer to these as pattern I (FoxA1 replicate 1),
II (FoxA1 replicate 2), and III (FoxA1 replicate 3), respectively.
Pattern III where the two arms along ARC are not distinguishable can
arise due to either low-complexity library or high sequencing depth.
For example, all three replicates of the TBP ChIP-exo from K562, with
sequencing depths between $\sim$ 60M to 115M reads, and replicate two
of TBP ChIP-nexus in K562, with sequencing depth of $\sim$ 130M reads,
exhibit this pattern. A simple but effective strategy to distinguish
the two plausible scenarios from Pattern III is to apply the QC
pipeline to sub-samples randomly generated from the full dataset at
varying sequencing depths. We applied this strategy by sub-sampling
20M to 50M reads, a range that represents the sequencing depths of the
human samples we are using in the paper, from the TBP samples.  URC
vs. ARC diagnostics of these sub-samples (Supplementary Figures S23 to
\ref{sfig:qcsamp3}) indicate that of the four TBP samples with this
pattern, replicates two and three of K562 ChIP-exo suffer from
low-complexity library issues, whereas the other samples exhibit the
pattern specific to high quality samples. \RW{To confirm this, we
  compared the top FIMO scores \cite{fimo} of the TBP motif calculated
  with the ChIP-exo and ChIP-nexus replicates. Figure~\ref{fig:5}E
  illustrates that the scores of the ChIP-nexus replicates and the
  first ChIP-exo replicate are higher than the other two samples when
  considering the top scores. As expected the ChIP-nexus scores are
  comparable among and show better performance than the three ChIP-exo
  experiment and among those three replicates, replicate 1 outperforms
  the other two.}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth,page = 1]{figures/fig3/Carroll_FoxA1_mouse_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page = 3]{figures/fig3/Carroll_FoxA1_mouse_strand_imbalance.pdf} 
  \includegraphics[width = .8\textwidth,page = 1]{figures/fig3/Carroll_FoxA1_mouse_strand_imbalance.pdf} 
  \caption{\RW{Diagnostic plots generated by \pname{}. Comparison of
      A) ARC vs. URC plot, B) Region Composition plot and C) FSR
      distribution plot between the three FoxA1 replicates in mouse
      liver tissue from \cite{exoillumina}.}}
  \label{fig:3}
\end{figure}

>>>>

The Strand Cross-Correlation (SCC), introduced by \cite{strandcc}, is the most commonly used quality metric in assessing ChIP-seq enrichment quality. It aims to quantify  how well the reads mapped to each strand are clustered around the locations of the protein-DNA interaction sites by calculating the Pearson correlation among forward and backward strands reads by shifting them across a range that covers both the read length of the experiment and the expected average fragment length. Typical SCC profiles exhibit two local maxima: at the average fragment length and the read length. In high quality experiments with clear ChIP enrichment, the average fragment length maximum coincides with the global maximum. In an idealized ChIP-exo experiment where the DNA fragments are digested to the boundaries of the protein-DNA interaction sites, we would expect the SCC profile to maximize at the motif length indicating clustering of the forward and reverse strand reads around the binding site. Figure~\ref{fig:1}D displays the SCC curves for the CTCF HeLa samples where the ChIP-exo curve shows local maxima at the motif and read lengths, while the SE ChIP-seq curves have a local maxima at the read length and a global maxima at the average fragment length. SCC profiles for other samples are available in Supplementary Figures \ref{sfig:scc1} to \ref{sfig:scc8}. The read length and motif length maxima are often in close proximity of each other; as a result, this renders QC metrics such as the Normalized Strand Cross-Correlation (NSC) or the Relative Strand Cross-Correlation (RSC) harder to interpret; however, the profile itself seems informative about the enrichment signal in ChIP-exo/nexus experiments. 
% \SK{this seems true at least based on my initial look at the plots, but we should be cautious in reporting this}. 
% \RW{i agree, and additionally Carroll et al, 2014 kinda agrees with us that the SCC-related measure may not be useful since the fragment length coincides with read length}
% http://journal.frontiersin.org/article/10.3389/fgene.2014.00075/full#h5 the paragraph is a little bit above conclusions 
% We continued our exploration by investigating whether the current state-of-the-art QC pipelines for ChIP-Seq are suitable for ChIP-exo %and ChIP-nexus. In Tables \ref{tab:qc_sig} and \ref{tab:qc} we calculated a collection of the commonly used ChIP-Seq QC metrics using the ChIP-exo and ChIP-nexus experiments instead: Normalized Strand Cross-Correlation (NSC), Relative Strand Cross-Correlation and PCR Bottleneck Coefficient (PBC) defined as in \url{https://genome.ucsc.edu/ENCODE/qualityMetrics.html}\nocite{encode_qc}.
% considerable amounts of reads are being mapped to specific positions.
% because of the exonuclease digestion, several reads that their $5\prime$ end was digested to the same position before the amplification step.
%, which by following the ChIP-Seq guidelines it would indicate that those experiment show severe bottlenecking problems, which may incorrectly suggest that the positions with a large amount of aligned reads are being caused by PCR amplification rather than observed genomic signal.
%, and usually it is expected to observed two local maxima, one when the profiles are shifted by the average read length and another when the profiles are shifted by the unobserved fragment length. In a high quality ChIP-Seq dataset the last one is also the SCC global maxima.

@@

To address the limitations of available analytical exploration approaches discussed above, we developed \pname{}. We first present the overall pipeline and then discuss individual components with a case study using ChIP-exo data of FoxA1 from \cite{exoillumina} and ChIP-nexus data from \cite{chipnexus}. Figure~\ref{fig:2} summarizes the 4-step pipeline.  Given aligned reads from a ChIP-exo/nexus sample, the first step partitions the reference genome into islands by keeping the non-digested ChIP-exo regions. In step 2, the total number of reads overlapping each island ($D_i$) and the number of island positions with at least one aligned read are recorded $(U_i)$. Then, three summary statistics $\text{ARC}_i$, $\text{URC}_i$, and $\text{FSR}_i$ are computed for each region $i$. $\text{ARC}_i$ denotes the \textit{average read coefficient} and is defined as the ratio of the \# of reads in island $i$ ($D_i$) to the width of the island $i$ ($W_i$); $\text{URC}_i$, \textit{unique read coefficient}, quantifies the inverse of the effective coverage and is defined as the ratio of the \# of genomic positions with at least one aligned read within island $i$ ($U_i$) to the \# of reads in island $i$ ($D_i$); and $\text{FSR}_i$ denotes the proportion of forward strand reads. Step 3 of the pipeline generates several diagnostic plots aimed at quantifying ChIP enrichment and strand imbalance and step 4 generates quantitative summaries of these diagnostic plots.

% (extended by the experiment's read length)
% \SK{check: extended or not? What do we extend to in ChIP-exo?}
% \RW{right now, we don't extend. we use the coverage plot with the experiment's read length to partition the coverage into the regions. In the package I left it as a parameter but I don't is going to change a lot. For example, the ChIP-exo datasets read length is ~45 and it's regions seems to be composed by scattered reads in both sides. So, I think that using a higher value   will only made the regions wider but will not increase the number of unique position or the depth by a lot}
%\begin{align*}
%  \text{ARC}_i &= \dfrac{\text{\#  of reads in island $i$}}{\text{Width of the island $i$}}, \\
%  \text{URC}_i & = \dfrac{\text{\# of genomic positions with at least
%      one aligned read within island $i$}}{\text{\# of reads in
%      island %$i$}}, \\
%      \text{FSR}_i &= \dfrac{\text{\# of forward strand reads in
%          island $i$}}{\text{\# of reads in island $i$}}.
%\end{align*}
Figure~\ref{fig:2}A presents the typical behaviour of the URC vs. ARC plot for a high quality ChIP-exo sample. In general, the plot depicts two strong arms. High URC values correspond to regions with reads concentrated on a small number of positions.  The left arm, with low ARC and varying URC values, corresponds to background islands, regions that are usually composed of scattered reads that were not digested during the exonuclease step. The right arm where the URC decreases as the ARC increases corresponds to regions that are usually ChIP enriched. We quantify the shape of the ARC vs. URC plot by the use of two estimated parameters: $\beta_1$ which represents the average number of reads aligned to the unique positions in large depth regions and $\beta_2$ which represents the overall change in depth as the width varies across a large set of regions. These parameters are estimated by sampling experiments on the original samples. We provide further details on them later in the paper where we apply the pipeline to a large collection of ChIP-exo/nexus experiments. Figures~\ref{fig:2}B and \ref{fig:2}C   present the typical behaviour of the \textit{Region Composition} and \textit{Forward Strand Ratio distribution} plots, both quantify the strand imbalance as part of the QC pipeline respectively, the former depicts how quickly the islands exclusively composed by fragments on a single strand are filtered out as islands with higher depths are observed. In a high quality sample, the proportion of islands with reads from only one strand is expected to decrease rapidly as we consider higher depth regions. In contrast, this proportion remains approximately constant in lower quality samples. The latter plot illustrates how quickly the quantiles of the FSR approaches to 0.5, the expected FSR value in high quality samples. Even though not every region in a ChIP-exo experiment is perfectly balanced, the most enriched regions are expected to be composed by approximately the same quantity of reads in both strands.

%and as the URCR decreases the library complexity does it as well, on the other hand 
% \SK{Not sure what this next sentence is referring to: Finally we
%   quantify the relationship between library complexity by the use of
%   two indexes that represent the change in the number of unique
%   positions per regions and the change in of the width of a regions as
%   the depth changes.}
%median, since in a high quality sample it is
%expected for the median to be approximately 0.5 and the enriched
%regions are going to be composed by fragments sequenced from both
%strands.
\subsubsection*{Application and validation of \pname{} with the FoxA1
  ChIP-exo dataset}
% \SK{We should present the whole QC results and end with validation -  panels from the two figures should be switched.}

We next use FoxA1 ChIP-exo datasets, with three biological replicates at comparable sequencing depths from mouse liver cells, to illustrate the QC pipeline.  Figure~\ref{fig:3}A presents URC vs. ARC plots for all three replicates. The first and third replicates exhibit a defined decreasing trend in URC as the ARC increases. This indicates that these samples exhibit a higher ChIP enrichment than the second replicate. On the other hand, the overall URC level from the first two replicates is higher than that of the third replicate level,elucidating that the libraries for the first two replicates are more complex than that of the third replicate.
%\SK{I have a hard time interpreting B and C specifically, need to include something about their implications for replicates 2 and 3.}

Figure~\ref{fig:3}B and \ref{fig:3}C display the Read Composition and FSR distribution plots, which highlight specific problems with replicates 2 and 3. Figure~\ref{fig:3}B, exhibits apparent decreasing trends in the proportions of regions formed by fragments in one exclusive strand. High quality experiments tend to show exponential decay in the proportion of single   stranded regions, while for the lower quality experiments, the trend may be linear or even constant (Supplement   Figure~\ref{sfig:qcsamp3}). FSR distributions of both of replicates 2 and 3 are more spread around their respective medians (Figure~\ref{fig:3}C). The approximation of the 0.1 and 0..9 quantiles indicate the aforementioned lower enrichment in the   second replicate and the low complexity in the third one.

Overall, we conclude that replicate 1 is higher quality than both of replicates 2 and 3.  We validate this observation with a motif analysis on the candidate binding regions identified from these replicates. A conservative approach to identify high quality binding regions (Materials and Methods) identify 7014, 1855 and 2187 regions for replicates 1, 2 and 3, respectively. The lower number of enriched regions from replicate 2 is consistent with the lower ChIP enrichment pattern in the $UCR$ vs. $ARC$ diagnostic plot. Scanning of these regions for the occurrence of FoxA1 sequence motif with the FIMO tool \cite{fimo} indicates that the first replicate outperforms the other two in terms of percentage of candidate regions with the FoxA1 motif  (Figure~\ref{fig:4}A).
% \SK{Could we say binding events? or is the peak analysis on the whole binding region, i.e., peak? If so, we may want o    change "candidate regions" to peaks?} 
% \RW{I used a set of high quality regions, i.e. using the ChIP-exo pipeline I filtered them by npos, FSR and depth. Figure~\ref{sfig:venn} show the overlap between these regions and the peaks (roughly speaking these regions are almost a proper subset of the peaks)}.

Furthermore, Figure~\ref{fig:4}B displays the average normalized read coverage around the actual motif locations in the candidate binding regions. These coverage plots reveal that the ChIP signal is more defined for the first and third replicates than the second one, indicating overall strength of the ChIP enrichment in these samples compared to the second replicate. Figure~\ref{fig:4}C compares the FIMO scores among the three replicates, not-surprisingly confirming that the first replicate exhibits the highest quality.

%As we discussed above,  background fragments are often digested by the exonuclease enzyme in ChIP-exo experiments; therefore, the balance between the enrichment and library complexity of an experiment is one of the key factors determining the sample's data quality. Using the Fox A1 in mouse liver cell lines generated by Serandour et al., 2013\nocite{exoillumina} and these two quantities, we explored the relationship between library complexity and experiment enrichment. 

@@

%The strand imbalance assessment is based in the observation that the enriched regions usually are composed of a higher quantity of reads, therefore we examined the FSR as the regions with lower depth are being filtered out. This indicator is of particular importance, as it evaluates the ``peak pair'' assumption that the original ChIP-exo paper suggested and multiple ChIP-exo data analysis methods rely on. For every ChIP-exo experiment, we calculated the global FSR and noticed that for all experiments is roughly $0.5$, which means there are approximately the same amount of reads in both strands.
% In particular, \RW{both replicate 2 and replicate 3 FSR's distributions are more spread around the median. The approximation of the 0.1 and 0.9 quantiles to the median is indicative of the aforementioned lower enrichment of the second replicate, since there are less regions with a higher concentration of reads, this makes the FSR's distribution to be dominated by non-enriched regions. On the other hand, the slow approximation of the quantiles to the FSR median in the third replicate is indicative of low library complexity. In Figure~\ref{fig:3}C, we can observe an decreasing trend in the proportion of regions formed by fragments in one exclusive strand. In a high quality experiment it is expected to observe an exponential trend, while in lower quality experiments the trend may be linear or even constant (Supplement Figure~\ref{sfig:qcsamp3}).}
%In order to study the strand imbalance of the FoxA1 factor ChIP-exo experiments in mouse liver generated by Serandour et al., %2013\nocite{exoillumina} we considered the visualizations shown in Figure \ref{fig:strand}: Figure \ref{fig:strand}A presents the FSR's behavior as the lower depth regions are being filtered out, while Figure \ref{fig:strand}B) shows the proportion of regions that are composed by reads from both strands against the regions formed by reads in exclusively one strand. In a high quality dataset, it is expected for all quantiles to quickly converging towards the median (in panel A) or the regions formed by reads in one strand (either forward or backward) to be formed by few fragments (in panel B). 
% Additionally, it is worth mentioning that the background in a ChIP-exo experiment is expected to be formed by the undigested ChIP fragments that are not bound to a protein, hence it is for the reads in both strand to be imbalanced in a background region. 
% Figure~\ref{fig:4}A summarizes the total numbers of regions identified from each  replicate. 
%Considering the regions and summary statistics we obtained a set of candidate sites for each replicate, the total number of candidate sites is shown in Figure \ref{fig:enrich}B. For this sites, we extracted the sequence and searched for the FoxA1 motif using \textbf{FIMO} \cite{fimo}. Figure \ref{fig:enrich}C shows the fraction of candidate site where a motif was found. The first replicate outperforms the rest in both number of candidate sites and proportion of sites with motif; on the other hand we can see that the second replicate candidate sites are more likely to contain the motif sequence, but there more candidate sites in the third replicate than in the second. 
% however,  overall, the signal pattern is reproducible in all of the three biological replicates.
% Figure~\ref{fig:4}D further highlights the overall quality of the identified motif sequences for each replicate and suggest that libraries with high library complexity (replicates 1 and \RW{2}) capture binding sites with better motif matches.
% Finally, Figure~\ref{fig:4}F compares the Strand Imbalance by considering the regions overlapping with their respective sets of ChIP-exo peaks. Not surprisingly, for the replicate 2 we can see that the Imbalance distribution differ among both classes, which is caused by the aforementioned lower enrichment of that sample; for the remaining replicates we can see that the Imbalance is more similar when regions composed by more reads are considered. This suggests that enriched regions are more  likely to be balanced, therefore this suggest that the ``peak-pair'' assumption may not hold in non-enriched regions.
%In summary,  the diagnostics plots of the QC pipeline highlights that  observing a defined decreasing trend in $\mbox{URCR}$ as $\mbox{ARC}$ increases implies ChIP enrichment and multiple samples can be compared with respect to the degree of this trend. Second, for experiments that are not deeply sequenced, low $\mbox{URCR}$ values represent low complexity. decreases the amount of sites of motif found but it doesn't modify the read distribution around the motif. Finally, libraries with higher complexity returns sequences with a better motif match. \SK{We do not have enough analysis with motif validation to conclude these.}
\subsubsection*{High sequencing depth may confound low-complexity library issues}

We next evaluated every sample listed in Tables~\ref{tab:qc} and
\ref{tab:qc_sig} with the \pname{} QC pipeline (Supplementary
Figures~\ref{sfig:qc1} to \ref{sfig:qc11}). A key, albeit not
surprising, observation from large scale analysis is that the URC
vs. ARC plots typically display the three patterns captured in the
FoxA1 study. We will refer to these as pattern I (FoxA1 replicate 1),
II (FoxA1 replicate 2), and III (FoxA1 replicate 3), respectively.
Pattern III where the two arms along ARC are not distinguishable can
arise due to either low-complexity library or high sequencing depth.
For example, all three replicates of the TBP ChIP-exo from K562, with
sequencing depths between $\sim$ 60M to 115M reads, and replicate two
of TBP ChIP-nexus in K562, with sequencing depth of $\sim$ 130M reads,
exhibit this pattern. A simple but effective strategy to distinguish
the two plausible scenarios from Pattern III is to apply the QC
pipeline to sub-samples randomly generated from the full dataset at
varying sequencing depths. We applied this strategy by sub-sampling
20M to 50M reads, a range that represents the sequencing depths of the
human samples we are using in the paper, from the TBP samples.  URC
vs. ARC diagnostics of these sub-samples (Supplementary Figures S23 to
\ref{sfig:qcsamp3}) indicate that of the four TBP samples with this
pattern, replicates two and three of K562 ChIP-exo suffer from
low-complexity library issues, whereas the other samples exhibit the
pattern specific to high quality samples. \RW{To confirm this, we
  compared the top FIMO scores \cite{fimo} of the TBP motif calculated
  with the ChIP-exo and ChIP-nexus replicates. Figure~\ref{fig:5}E
  illustrates that the scores of the ChIP-nexus replicates and the
  first ChIP-exo replicate are higher than the other two samples when
  considering the top scores. As expected the ChIP-nexus scores are
  comparable among and show better performance than the three ChIP-exo
  experiment and among those three replicates, replicate 1 outperforms
  the other two.}

@@

\begin{figure}[h!]
  \centering
%  \includegraphics[width = .8\textwidth,page = 1]{figures/fig3/Carroll_FoxA1_mouse_enrichment.pdf}
%  \includegraphics[width = .8\textwidth,page = 3]{figures/fig3/Carroll_FoxA1_mouse_strand_imbalance.pdf} 
%  \includegraphics[width = .8\textwidth,page = 1]{figures/fig3/Carroll_FoxA1_mouse_strand_imbalance.pdf} 
    \includegraphics[width = .95\textwidth]{figures/fig3/fig3.pdf}
  \caption{\RW{Diagnostic plots generated by \pname{}. Comparison of
      A) ARC vs. URC plot, B) Region Composition plot and C) FSR
      distribution plot between the three FoxA1 replicates in mouse
      liver tissue from \cite{exoillumina}.}}
  \label{fig:3}
\end{figure}
